Comparison of clusters to cell types for `r params$library_id`

### 2024-11-21 analyses/cell-type-ewings/template_notebooks/clustering-workflow/02-clustering-celltypes.Rmd

---
title: "Clustering exploration for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: NULL
  sce_file: NULL
  singler_results_file: NULL
  cluster_results_file: NULL
  marker_genes_file: NULL
---

This notebook summarizes clustering for `r params$library_id`.

- Louvain, Jaccard clusters are calculated using a range of parameters specifying nearest neighbors (5, 10, 15, 20, 25, 30, 35, and 40). 
- Metrics are then calculated to evaluate clustering results across all parameters: 
  - Silhouette width 
  - Cluster purity 
  - Cluster stability 
- The clusters are then compared to the results from running `SingleR` in the `aucell-singler-annotation.sh` workflow.
- We then look at marker gene expression across the clusters and assign each cluster to a cell type.

@@

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)
```

@@

# read in singler results
singler_classification_df <- readr::read_tsv(params$singler_results_file)
```

## Clustering

Below we perform Louvain, Jaccard clustering, varying `k`. 
The minimum `k` is 5 and the maximum `k` is 40 with a step size of 5.

```{r}
# perform clustering across k = 5-40 with increments of 5
all_cluster_results <- cluster_sweep(sce) |>
  dplyr::mutate(
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

```{r}
# get umap embeddings and combine into a data frame with cluster assignments
umap_df <- sce |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1 and get rid of excess columns
  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
  dplyr::left_join(all_cluster_results, by = "barcodes")
```

Below we visualize the cluster assignments using each parameter on a UMAP. 
This can be helpful to see any values of `k` that may have obvious over or under clustering.

```{r, fig.height=15, fig.width=10}
# look at clustering results for each library across all params
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(alpha = 0.5, size = 0.1) +
  facet_wrap(vars(nn_param)) +
  theme(
    aspect.ratio = 1,
    legend.position = "none"
  )
```

## Clustering statistics

Below we calculate a series of statistics:

- Average silhouette width: This metric evaluates cluster separation. 
Cells with large positive silhouette widths are closer to other cells in the same cluster than to cells in different clusters. 
Higher values indicate tighter clusters.
- Average cluster purity: This metric also evaluates cluster separation and tells us the proportion of neighboring cells that are assigned to the same cluster. 
Purity values range from 0-1 with higher purity values indicating clusters that are well separated. 
- Cluster stability: This evaluates how stable the clustering is to input data. 
Stability values range from 01- with higher values of cluster stability indicating more reproducible clusters.

```{r, warning=FALSE}
# get a combined stats dataframe with purity and width for all clusters
all_stats_df <- get_cluster_stats(sce, all_cluster_results) |>
  dplyr::mutate(
    # make sure the order is correct
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

### Silhouette width

```{r}
# silhouette width for different params
plot_cluster_stats(all_stats_df, width)
```

### Cluster purity

```{r}
# cluster purity for different params
plot_cluster_stats(all_stats_df, purity)
```

### Cluster stability

```{r, warning=FALSE, message=FALSE}
# calculate cluster stability
stability_df <- get_cluster_stability(sce, all_cluster_results) |>
  dplyr::mutate(
    k_value = as.factor(k_value),
    # make sure that k = 5 comes first
    k_value = forcats::fct_relevel(k_value, "5", after = 0)
  )

# plot stability across all values of k
ggplot(stability_df, aes(x = k_value, y = ari)) +
  geom_jitter(width = 0.1) +
  labs(title = "Cluster stability") +
  stat_summary(
    aes(group = k_value),
    color = "red",
    # median and quartiles for point range
    fun = "median",
    fun.min = function(x) {
      quantile(x, 0.25)
    },
    fun.max = function(x) {
      quantile(x, 0.75)
    }
  )
```

## Compare clusters to cell types from `SingleR`

@@

```{r}
cluster_classification_df <- all_cluster_results |>
  # add in classifications from singler
  dplyr::left_join(singler_classification_df, by = "barcodes") |>
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf),
    nn_param = forcats::fct_relevel(nn_param, "k_5", after = 0)
  )
```

@@

## Marker gene expression across clusters

>>>>

---
title: "Comparison of clusters to cell types for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: "SCPCL000822"
  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
  singler_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_singler-classifications.tsv"
  cluster_results_file: "../../results/cluster_exploration/SCPCS000490/SCPCL000822_clusters.tsv"
  marker_genes_file: "../../references/visser-all-marker-genes.tsv"
---

TODO: Update this notebook to take as input a clustering algorithm, nn (range of nn?), resolution, and optional objective function to display in SingleR and marker gene plots

- The clusters are then compared to the results from running `SingleR` in the `aucell-singler-annotation.sh` workflow.
- We then look at marker gene expression across the clusters and assign each cluster to a cell type.

@@

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)

# set seed
set.seed(2024)
```

@@

# read in singler results
singler_classification_df <- readr::read_tsv(params$singler_results_file) |>
  dplyr::rename("cell_id" = "barcodes")
```

## Clusters vs. `SingleR` {.tabset}

@@

```{r}
cluster_classification_df <- all_cluster_results |>
  # add in classifications from singler
  dplyr::left_join(singler_classification_df, by = "cell_id") |>
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf)
  )

cluster_classification_df |> 
  split(cluster_classification_df$resolution) |> 
  purrr::map(\(df){cluster_celltype_heatmap(df)})

```

@@

## Marker gene expression {.tabset}
